Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-09-28 main_new.tex
Language input - Ginny

Data is gathered by an AUV that can measure temperature and salinity
at a subset of locations on the defined grid. This can be done in a
static manner (where the survey design is pre-planned), or in a
sequential way with several rounds of data gathering. For the latter,
a round of data is gathered first, and then assimilated to get an
updated model. Then the updated model is used to design and select the
next round of measurements, and so on. We describe the model for one
round of data and the Gaussian formula for a single update. In Section
\ref{sec:heuristics} we discuss sequential data gathering in the
context of adaptive design of AUV measurements.

@@

\subsection{Excursion probabilities for Gaussian processes}

Based on Gaussian modeling assumptions, the EPs can be computed at any
location $\bx$. Without additional data, Eq.  (\ref{eq:prob}) gives us:

@@

When data $\by$ are available, Eq. (\ref{eq:post_ep}) similarly gives
us:
\begin{eqnarray}\label{eq:prob_mv}
 p(\bx;\by) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by)
 \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} t_T\\
t_S
\end{bmatrix};
\begin{bmatrix} m_{T}(\bx)\\
m_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
a_{T}^2(\bx) & a_{T,S}(\bx)  \\
a_{T,S}(\bx)  & a_{S}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
with parameters as defined in Eq. (\ref{gp_upd}) and (\ref{gp_hat}). Here, the probabilities would not only depend on the data but also on the data design defined via matrix $\bG$ in Eq. (\ref{likelihood}). We next derive results that ease comparison of several designs.

@@

We will next derive closed form solutions to the expectation in
(\ref{sur}). The location index $\bx$ is suppressed to avoid overly
complex notation. We present new results for computing the expectation
$E_{\by}\left\{ \cdot \right\}$ which is the integrand of
Eq. (\ref{sur}).

A critical first element in the derivation is using the GP model and the
linear dependence on $\by$ in the conditional mean in
Eq. (\ref{gp_upd}). This results in a closed form Gaussian
distribution for the conditional mean in Eq. (\ref{dist_mxi}). The
high-dimensional inner integral in Eq. (\ref{sur}) then reduces to a
bivariate integral \citep{bhattacharjya2013value, chevalier2014fast},
with expectation over $\bm=(m_{T},m_{S})^t$.  We standardize the two
variables in Eq. (\ref{sur}), i.e.  $Z_T=\frac{\xi_T-m_{T}}{a_{T}}$,
$Z_S=\frac{\xi_S-m_{S}}{a_{S}}$, and then
\begin{eqnarray}\label{ytom}
   P(\xi_T \leq t_T, \xi_S \leq t_S|\by) &=& P \left( Z_T \leq\frac{t_T-m_{T}}{a_{T}}, Z_S \leq\frac{t_S-m_{S}}{a_{S}}|\by \right) \nonumber \\
   &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} \frac{t_T-m_{T}}{a_{T}}\\
\frac{t_S-m_{S}}{a_{S}}
\end{bmatrix};
 \begin{bmatrix} 0\\
0
\end{bmatrix},\begin{bmatrix}
1 & \eta_{s,t}  \\
\eta_{s,t}   & 1  
\end{bmatrix}\end{pmatrix} \nonumber \\
&=& P(\xi_T \leq t_T, \xi_S \leq t_S|\bm),
\end{eqnarray}
where $\eta_{T,S} =a_{T,S}/(a_{T} a_{S})$.
Since the expression only depends on $\by$ via its affine function $\bm$, the expectation in Eq. (\ref{sur}) is reduced to a bivariate integral over $\bm$. 
Introducing $\pi(\bm)$ for the density
function of $\bm$, the expected Bernoulli variance becomes:
\begin{eqnarray}\label{eq:var2}
E_{\by}(p(1-p)) &=& E_{\bm}(p(1-p)), \hspace{3mm} p=P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) \nonumber \\
E_{\bm}(p(1-p)) & = & \int p[1-p] \pi(\bm) d\bm, \nonumber \\
 &=& \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm)  \pi(\bm) d\bm \nonumber  \\
&-& \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) \pi(\bm) d\bm. 
\end{eqnarray}
We will next extend results of \cite{chevalier2014fast} to solve for
these two parts (see also \cite{stroh} for such an extensions).

For the first part of the above, we have:
\begin{equation}\label{part1:phi2}
 \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) \pi(\bm) d\bm= 
P \left( Z_{T} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{S} \leq \frac{t_S-m_{S}}{a_{S}} \right), \nonumber
\end{equation}
where $\bZ=(Z_{T},Z_{S})$ is a bivariate zero-mean and unit-variance
Gaussian vector, with $\mbox{Corr}(Z_{T},Z_{S})=\eta_{T,S}$. Moreover,
$\bZ$ is chosen to be independent of $\bm$. Using Eq. (\ref{dist_mxi}) we have
\begin{equation}
    E(a_{T} Z_{T}+m_{T}-t_T) = \mu_{T}-t_T, \hspace{3mm}
    \mbox{Var}(a_{T} Z_{T}+m_{T}-t_T) = a^2_T+ \psi^2_{T}, 
\end{equation}
for the temperature part and the same holds for the salinity part. From this we get:
\begin{eqnarray}\label{two_parts0}
& P & \left( Z_{T} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{S} \leq \frac{t_S-m_{S}}{a_{S}} \right) \\
&=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} 0\\
0
\end{bmatrix};
\begin{bmatrix} \mu_{T}-t_T\\
\mu_{S}-t_S
\end{bmatrix},\begin{bmatrix}
a^2_T+ \psi^2_{T} & a_{T,S}+\psi_{T,S}  \\
a_{T,S}+\psi_{T,S}   & a^2_T+ \psi^2_{T} 
\end{bmatrix}\end{pmatrix} \nonumber.
\end{eqnarray}

For the second part of the expression (\ref{eq:var2}), standardization
implies that: 
\begin{eqnarray}\label{part1:phi4}
&& \hspace{6mm} \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) p(\bm) d\bm =  \\
&P& \hspace{-3mm} \left( Z_{1,T} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{1,S} \leq \frac{t_S-m_{S}}{a_{S}},Z_{2,S} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{2,S} \leq \frac{t_S-m_{S}}{a_{S}} \right). \nonumber
\end{eqnarray}
Here, $\bZ_1=(Z_{1,T},Z_{1,S})$, $\bZ_2=(Z_{2,T},Z_{2,S})$ are
independent bivariate zero-mean and unit-variance Gaussian vectors,
both with element-wise correlation $\eta_{T,S}$. Moreover, $\bZ_1$ and
$\bZ_2$ are chosen to be independent of $\bm$. Hence:
\begin{equation}
    \left(
    \begin{array}{cccccc}
         Z_{1,T} \\
         Z_{1,S} \\
         Z_{2,T} \\
         Z_{2,S} \\
         m_{T} \\
         m_{S}
    \end{array}
    \right)
\sim N \left(
\left(
    \begin{array}{cccccc}
    0 \\
    0 \\
    0 \\
    0 \\
          \mu_{T} \\
         \mu_{S} 
    \end{array}
    \right),
   \left[ 
      \begin{array}{cccccc}
        1 & \eta_{T,S} & 0 & 0 & 0 & 0  \\
        \eta_{T,S} & 1 & 0 & 0 & 0 & 0 \\
        0 & 0 & 1 & \eta_{T,S} & 0 & 0 \\
        0 & 0 & \eta_{T,S} & 1 & 0 & 0 \\
        0 & 0 & 0 & 0 & \psi^2_{T} & \psi_{T,S} \\
        0 & 0 & 0 & 0 & \psi_{T,S} & \psi^2_{S} 
    \end{array}
\right]
\right).
\vspace{0.2cm}
\end{equation}

The solution to the second part of the equation above, is then to
evaluate a four-variable cumulative distribution function associated
with Eqn. (\ref{part1:phi4}), accounting for the mean, variance and
covariances of $\bZ_1$, $\bZ_2$ and $\bm$ in the appropriate linear
combinations. Using vector-matrix formulations, we define linear
combinations:
\begin{equation}\label{Vcomb}
    \bV=\left[
    \begin{array}{cccccc}
        a_T & 0 & 0 & 0 & 1 & 0 \\
         0 & a_S & 0 & 0 & 0 & 1 \\
         0 & 0 & a_T & 0 & 1 & 0\\
         0 & 0 & 0 & a_S & 0 & 1\\
    \end{array}
    \right] 
    \left(
    \begin{array}{cccccc}
         Z_{1,T} \\
         Z_{1,S} \\
         Z_{2,T} \\
         Z_{2,S} \\
         m_{T} \\
         m_{S}
    \end{array}
    \right),
\end{equation}
and this in turn then becomes:
\begin{eqnarray}\label{part1:phi4_2}
&& P \left( Z_{1,T} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{1,S} \leq \frac{t_S-m_{S}}{a_{S}},Z_{2,S} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{2,S} \leq \frac{t_S-m_{S}}{a_{S}} \right)  \\
&=&  \Phi_4 
\begin{pmatrix} 
\begin{bmatrix} t_T\\
t_S \\
t_T \\
t_S
\end{bmatrix};
E( \bV),Var(\bV)
\end{pmatrix} \nonumber.
\end{eqnarray}

The multivariate cumulative probabilities in two and four dimensions
are fast to compute using methods of \cite{genz2009computation}.  In
the end, the first part in Eq. (\ref{two_parts0}) and the second in
Eq. (\ref{part1:phi4}) are computed for each location on the grid, and
summed over $\bx \in \mathcal{M}_g$, gives the design evaluation
criterion in Eq. (\ref{sur}).

@@

In our application domain with the variables temperature and salinity,
the solutions to the complex integral equations for the expected IBV
are reduced to calculating bi-variate and four-variate cumulative
distribution functions for Gaussian vectors. For the more general
situation one would be interested in $K$ random fields. Say, in the
ocean sciences the variables could include chlorophyll or other
bio-geochemical quantities, in addition to temperature and salinity.

The equations derived in the previous section can then be generalized. Let
$\bxi(\bx)=(\xi_1(\bx),\ldots,\xi_K(\bx))$, and the expected Bernoulli
variance is:
\begin{equation}\label{two_partsK}
E_{\by}(p(\bx;\by)(1-p(\bx;\by))), \hspace{3mm} p(\bx;\by)=P(\xi_k(\bx) \leq t_k ; k=1,\ldots,K|\by). 
\end{equation}
By the first key result of linear combination of Gaussian variables,
we reduce the integral to a $K$ dimensional integral of the relevant
$\bm=E(\bxi(\bx)|\by)$. Next, reducing the integral to two parts, we
standardize the variables and compute cumulative distributions.  In
summary, we get:
\begin{equation}\label{two_parts}
E_{\by}(p(\bx)(1-p(\bx))) =  \Phi_K 
\begin{pmatrix}
\begin{bmatrix} t_1\\
\vdots \\
t_K 
\end{bmatrix};
\bmu,\bA+\bPsi 
\end{pmatrix}
- \Phi_{2K} 
\begin{pmatrix}
\begin{bmatrix} t_1\\
\vdots \\
t_K \\
t_1\\
\vdots \\
t_K 
\end{bmatrix};
E(\bV),Var(\bV) 
\end{pmatrix},\nonumber
\end{equation}
where $\bA$ and $\bPsi$ are direct $K$-variate extensions of the
matrices in Eq. (\ref{two_parts0}), and the matrix $\bV$ finds the
relevant linear combination of two independent vectors
$\bZ_l=(Z_{l,1},\ldots,Z_{l,K})$, $l=1,2$ and $\bm$, as an extension
of Eq. (\ref{Vcomb}).

\subsection{An Example}

We calculate EPs and the Bernoulli variance for a few bivariate
Gaussian distributions to illustrate the concepts articulated above.
Fig. \ref{illus_bivarDens} shows contour plots of three different
densities with increasing correlation $\gamma$ between temperature and
salinity.
\begin{figure}[h!] \centering
  \includegraphics[width=0.99\textwidth]{Figures/illus_bivar.eps}
  \caption{Density contour plots having different correlations between
    temperature and salinity. The densities have unit variance and the
    thresholds are identical to the mean values $5^o C$ and
    $30 mg/l$. x-axis is temperature and y-axis is salinity.}
\label{illus_bivarDens}
\end{figure}
Here, the thresholds are set equal to the mean, $\mu_T=t_T=5^o C$ and
salinity $\mu_S=t_S=30$ mg/l. The displayed densities have unit
standard deviations for both temperature and salinity, but we also
study the effect of doubling the standard deviations.

Table \ref{tab:sim_rhoab} shows the EPs and the associated Bernoulli
variance (second row) for the examples indicated in
Fig. \ref{illus_bivarDens}. The EPs increase with the correlation as
there is a stronger tendency of having jointly small temperature and
salinity. The Bernoulli variance is similarly largest for high
correlation. EPs and Bernoulli variances are the same for standard
deviation $1$ or $2$, and this implies that high variability in
temperature and salinity is not captured in the $p(1-p)$ expression.

\begin{table}[!h] \centering \caption{EP and Bernoulli variance for
    different correlations and variances (top rows), and expected
    Bernoulli variances for both temperature and salinity data $\by$ and only for
    temperature $y_T$ (bottom rows).}
  \begin{tabular}{c|ccc|ccc}
 &\multicolumn{3}{c}{$\sigma_S=\sigma_T=1$} & \multicolumn{3}{c}{$\sigma_S=\sigma_T=2$} \\
\hline
Correlation $\gamma$ & 0.2 & 0.6 & 0.8 & 0.2 & 0.6 & 0.8 \\
\hline
$p$ & 0.28 & 0.35 & 0.40 & 0.28 & 0.35 & 0.40 \\ 
$p(1-p)$ & 0.20 & 0.23 & 0.24 & 0.20 & 0.23 & 0.24 \\ 
$E_{\by}(p(\by) (1-p(\by)))$ & 0.092 & 0.089 & 0.085 & 0.052 & 0.051 & 0.049 \\ 
$E_{y_T}(p(y_T) (1-p(y_T)))$ & 0.151 & 0.138 & 0.123 & 0.137 & 0.114 & 0.093 \\ 
\hline
\end{tabular}
\label{tab:sim_rhoab}
\end{table}

Table \ref{tab:sim_rhoab} (bottom two rows) shows results of expected
Bernoulli variance calculations. This is presented for a design
gathering both data types $(y_T,y_S)$, and for a design with
temperature measurements $y_T$ alone. Having both data;
$(y_T,y_S)^t=(\xi_T,\xi_S)^t+N(0,0.5^2\bI)$, while
$y_T=\xi_T+N(0,0.5^2)$ when only temperature is measured.  For this illustration, Table \ref{tab:sim_rhoab} shows that the expected Bernoulli variance
gets lower with larger standard deviations $\sigma_T$ and $\sigma_S$
(right columns). The reduction of Bernoulli variance is largest
for the cases with high correlation $\gamma$. Albeit smaller, there is
also uncertainty reduction when only temperature is measured (bottom
row), especially when temperature and salinity are very
dependent. When correlation is low ($\gamma=0.2$) there is little
information about salinity in the temperature data, and hence less
uncertainty reduction. In an application with fresh cold water from
river source, the temperature and salinity variables will not only be
interdependent, but will show dependence in the spatial
dimension. This in turn will impact the design criteria when we
evaluate the information measure by integrating over several locations
$\bx$.

@@

An important synergy underlining this work is the potential of robotic sampling to adapt survey plans based on data and statistical inference,
in accordance with the sense-plan-act loop
(Fig. \ref{fig:sense-plan-act}). We will therefore present a few adaptive strategies
for robotic sampling that builds on the notion of multivariate ES. Simulations, analysis, and  results from full-scale trials are presented.

@@

An adaptive AUV survey is split in many rounds (stages), and at each
round, the survey design is found by using a myopic approach of
considering the largest reduction in uncertainty of the ES. In our
setting the selection is made on the graph defined by the nearest grid
nodes in the domain $\mathcal{M}_g$.

>>>>

Data is gathered by an AUV that can measure temperature and salinity
at a subset of locations on the defined grid. This can be done in a
static manner (where the survey design is pre-planned), or in a
sequential way with several rounds of data gathering. For the latter,
a round of data is gathered first, and then assimilated to get an
updated model. Then the updated model is used to design and select the
next round of measurements, and so on. Below, we describe the model for one round of data and the Gaussian formula for a single update; in Section \ref{sec:heuristics}, we discuss sequential data gathering in the context of adaptive design of AUV measurements.

@@

\subsection{Excursion Probabilities for Gaussian Processes}

Based on Gaussian modeling assumptions, the EPs can be computed at any
location $\bx$. Without additional data, Eq.  (\ref{eq:prob}) gives us

@@

When data $\by$ are available, Eq. (\ref{eq:post_ep}) similarly gives
us
\begin{eqnarray}\label{eq:prob_mv}
 p(\bx;\by) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by)
 \nonumber \\
 &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} t_T\\
t_S
\end{bmatrix};
\begin{bmatrix} m_{T}(\bx)\\
m_{S}(\bx)
\end{bmatrix},\begin{bmatrix}
a_{T}^2(\bx) & a_{T,S}(\bx)  \\
a_{T,S}(\bx)  & a_{S}^2(\bx)  
\end{bmatrix}\end{pmatrix},
\end{eqnarray}
with parameters as defined in Eq. (\ref{gp_upd}) and (\ref{gp_hat}). Here, the probabilities would not only depend on the data but also on the data design defined via matrix $\bG$ in Eq. (\ref{likelihood}). We next derive results that simplify the comparison of several designs.

@@

We will next derive closed-form solutions for the expectation in
(\ref{sur}). The location index $\bx$ is suppressed to avoid overly
complex notation. We present new results for computing the expectation
$E_{\by}\left\{ \cdot \right\}$ which is the integrand of
Eq. (\ref{sur}).

A critical first element in the derivation is using the GP model and the
linear dependence on $\by$ in the conditional mean in
Eq. (\ref{gp_upd}). This results in a closed form Gaussian
distribution for the conditional mean in Eq. (\ref{dist_mxi}). The
high-dimensional inner integral in Eq. (\ref{sur}) then reduces to a
bivariate integral \citep{bhattacharjya2013value, chevalier2014fast},
with an expectation over $\bm=(m_{T},m_{S})^t$.  We standardize the two
variables in Eq. (\ref{sur}) as   $Z_T=\frac{\xi_T-m_{T}}{a_{T}}$,
$Z_S=\frac{\xi_S-m_{S}}{a_{S}}$; we can then rephrase the probability as
\begin{eqnarray}\label{ytom}
   P(\xi_T \leq t_T, \xi_S \leq t_S|\by) &=& P \left( Z_T \leq\frac{t_T-m_{T}}{a_{T}}, Z_S \leq\frac{t_S-m_{S}}{a_{S}}|\by \right) \nonumber \\
   &=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} \frac{t_T-m_{T}}{a_{T}}\\
\frac{t_S-m_{S}}{a_{S}}
\end{bmatrix};
 \begin{bmatrix} 0\\
0
\end{bmatrix},\begin{bmatrix}
1 & \eta_{s,t}  \\
\eta_{s,t}   & 1  
\end{bmatrix}\end{pmatrix} \nonumber \\
&=& P(\xi_T \leq t_T, \xi_S \leq t_S|\bm),
\end{eqnarray}
where $\eta_{T,S} =a_{T,S}/(a_{T} a_{S})$.
Since the expression only depends on $\by$ via its affine function $\bm$, the expectation in Eq. (\ref{sur}) is reduced to a bivariate integral over $\bm$. 
Introducing $\pi(\bm)$ for the density
function of $\bm$, the expected Bernoulli variance becomes:
\begin{eqnarray}\label{eq:var2}
E_{\by}(p(1-p)) &=& E_{\bm}(p(1-p)), \hspace{3mm} p=P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) \nonumber \\
E_{\bm}(p(1-p)) & = & \int p[1-p] \pi(\bm) d\bm, \nonumber \\
 &=& \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm)  \pi(\bm) d\bm \nonumber  \\
&-& \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) \pi(\bm) d\bm. 
\end{eqnarray}
We will next extend results of \cite{chevalier2014fast} to solve for
these two parts (see also \cite{stroh} for such an extensions).

For the first part of the integral above, we have
\begin{equation}\label{part1:phi2}
 \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) \pi(\bm) d\bm= 
P \left( Z_{T} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{S} \leq \frac{t_S-m_{S}}{a_{S}} \right), \nonumber
\end{equation}
where $\bZ=(Z_{T},Z_{S})$ is a bivariate zero-mean and unit-variance
Gaussian vector, with $\mbox{Corr}(Z_{T},Z_{S})=\eta_{T,S}$. Moreover,
$\bZ$ is chosen to be independent of $\bm$. Using Eq. (\ref{dist_mxi}), we have
\begin{equation}
    E(a_{T} Z_{T}+m_{T}-t_T) = \mu_{T}-t_T, \hspace{3mm}
    \mbox{Var}(a_{T} Z_{T}+m_{T}-t_T) = a^2_T+ \psi^2_{T} 
\end{equation}
for the temperature part; the same holds for the salinity part. From this we get:
\begin{eqnarray}\label{two_parts0}
& P & \left( Z_{T} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{S} \leq \frac{t_S-m_{S}}{a_{S}} \right) \\
&=& \Phi_2 \begin{pmatrix} 
\begin{bmatrix} 0\\
0
\end{bmatrix};
\begin{bmatrix} \mu_{T}-t_T\\
\mu_{S}-t_S
\end{bmatrix},\begin{bmatrix}
a^2_T+ \psi^2_{T} & a_{T,S}+\psi_{T,S}  \\
a_{T,S}+\psi_{T,S}   & a^2_T+ \psi^2_{T} 
\end{bmatrix}\end{pmatrix} \nonumber.
\end{eqnarray}

For the second part of the expression (\ref{eq:var2}), standardization
implies that: 
\begin{eqnarray}\label{part1:phi4}
&& \hspace{6mm} \int P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) P(\xi_T \leq t_T, \xi_S \leq t_S|\bm) p(\bm) d\bm =  \\
&P& \hspace{-3mm} \left( Z_{1,T} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{1,S} \leq \frac{t_S-m_{S}}{a_{S}},Z_{2,S} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{2,S} \leq \frac{t_S-m_{S}}{a_{S}} \right). \nonumber
\end{eqnarray}
Here, $\bZ_1=(Z_{1,T},Z_{1,S})$, $\bZ_2=(Z_{2,T},Z_{2,S})$ are
independent bivariate zero-mean and unit-variance Gaussian vectors,
both with element-wise correlation $\eta_{T,S}$. Moreover, $\bZ_1$ and
$\bZ_2$ are chosen to be independent of $\bm$, hence:

\begin{equation}
\small
\label{eq:22}
    \left(
    \begin{array}{cccccc}
         Z_{1,T} \\
         Z_{1,S} \\
         Z_{2,T} \\
         Z_{2,S} \\
         m_{T} \\
         m_{S}
    \end{array}
    \right)
\sim N \left(
\left(
    \begin{array}{cccccc}
    0 \\
    0 \\
    0 \\
    0 \\
          \mu_{T} \\
         \mu_{S} 
    \end{array}
    \right),
   \left[ 
      \begin{array}{cccccc}
        1 & \eta_{T,S} & 0 & 0 & 0 & 0  \\
        \eta_{T,S} & 1 & 0 & 0 & 0 & 0 \\
        0 & 0 & 1 & \eta_{T,S} & 0 & 0 \\
        0 & 0 & \eta_{T,S} & 1 & 0 & 0 \\
        0 & 0 & 0 & 0 & \psi^2_{T} & \psi_{T,S} \\
        0 & 0 & 0 & 0 & \psi_{T,S} & \psi^2_{S} 
    \end{array}
\right]
\right).
\vspace{0.2cm}
\end{equation}

The solution to the second part of Eq. \eqref{eq:22} can be found by evaluating a four-variable cumulative distribution function associated
with Eq. (\ref{part1:phi4}), accounting for the mean, variance and
covariances of $\bZ_1$, $\bZ_2$ and $\bm$ in the appropriate linear
combinations. Using vector-matrix formulations, we define the linear
combinations
\begin{equation}
\label{Vcomb}
    \bV=\left[
    \begin{array}{cccccc}
        a_T & 0 & 0 & 0 & 1 & 0 \\
         0 & a_S & 0 & 0 & 0 & 1 \\
         0 & 0 & a_T & 0 & 1 & 0\\
         0 & 0 & 0 & a_S & 0 & 1\\
    \end{array}
    \right] 
    \left(
    \begin{array}{cccccc}
         Z_{1,T} \\
         Z_{1,S} \\
         Z_{2,T} \\
         Z_{2,S} \\
         m_{T} \\
         m_{S}
    \end{array}
    \right),
\end{equation}
which in turn then becomes:
{\small
\begin{eqnarray}
\label{part1:phi4_2}
&&P \left( Z_{1,T} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{1,S} \leq \frac{t_S-m_{S}}{a_{S}}, Z_{2,S} \leq \frac{t_T-m_{T}}{a_{T}}, Z_{2,S} \leq \frac{t_S-m_{S}}{a_{S}} \right)\\
&=&  \Phi_4 
\begin{pmatrix} 
\begin{bmatrix} t_T\\
t_S \\
t_T \\
t_S
\end{bmatrix};
E( \bV),Var(\bV)
\end{pmatrix} \nonumber.
\end{eqnarray}
}
The multivariate cumulative probabilities in two and four dimensions
are fast to compute using methods developed by \cite{genz2009computation}.  In
the end, the first part in Eq. (\ref{two_parts0}) and the second in
Eq. (\ref{part1:phi4}) are computed for each location on the grid and
summed over $\bx \in \mathcal{M}_g$, which gives the design evaluation
criterion in Eq. (\ref{sur}).

@@

In our application domain with the variables temperature and salinity,
the solutions to the complex integral equations for the expected IBV
are reduced to calculating bi-variate and four-variate cumulative
distribution functions for Gaussian vectors. For the more general
situation, one would be interested in $K$ random fields. In the
ocean sciences, the variables could include chlorophyll or other
bio-geochemical quantities in addition to temperature and salinity.

The equations derived in the previous section can then be generalized. Let
$\bxi(\bx)=(\xi_1(\bx),\ldots,\xi_K(\bx))$, and the expected Bernoulli
variance is:
\begin{equation}\label{two_partsK}
E_{\by}(p(\bx;\by)(1-p(\bx;\by))), \hspace{3mm} p(\bx;\by)=P(\xi_k(\bx) \leq t_k ; k=1,\ldots,K|\by). 
\end{equation}
From the first key result of linear combination of Gaussian variables,
we reduce the integral to a $K$ dimensional integral of the relevant
$\bm=E(\bxi(\bx)|\by)$. Next, reducing the integral to two parts, we
standardize the variables and compute cumulative distributions.  In
summary, we get
\begin{equation}\label{two_parts}
E_{\by}(p(\bx)(1-p(\bx))) =  \Phi_K 
\begin{pmatrix}
\begin{bmatrix} t_1\\
\vdots \\
t_K 
\end{bmatrix};
\bmu,\bA+\bPsi 
\end{pmatrix}
- \Phi_{2K} 
\begin{pmatrix}
\begin{bmatrix} t_1\\
\vdots \\
t_K \\
t_1\\
\vdots \\
t_K 
\end{bmatrix};
E(\bV),Var(\bV) 
\end{pmatrix},\nonumber
\end{equation}
where $\bA$ and $\bPsi$ are direct $K$-variate extensions of the
matrices in Eq. (\ref{two_parts0}), and the matrix $\bV$ finds the
relevant linear combination of two independent vectors
$\bZ_l=(Z_{l,1},\ldots,Z_{l,K})$, $l=1,2$ and $\bm$, as an extension
of Eq. (\ref{Vcomb}).

\subsection{An Example of IBV using Temperature and Salinity}

In this section, we calculate EPs and the Bernoulli variance for a few bivariate
Gaussian distributions to illustrate the concepts articulated above.
Fig. \ref{illus_bivarDens} shows contour plots of three different
densities with increasing correlation $\gamma$ between temperature and
salinity.
\begin{figure}[h!] \centering
  \includegraphics[width=0.99\textwidth]{Figures/illus_bivar.eps}
  \caption{Density contour plots with different correlations between
    temperature and salinity. The densities have unit variance and the
    thresholds are identical to the mean values $5^o C$ and
    $30 mg/l$. X-axis is temperature and y-axis is salinity.}
\label{illus_bivarDens}
\end{figure}
Here, the thresholds are set equal to the mean $\mu_T=t_T=5^o C$, and
salinity $\mu_S=t_S=30$ mg/l. The displayed densities have unit
standard deviations for both temperature and salinity, but we also
study the effect of doubling the standard deviations.

Table \ref{tab:sim_rhoab} shows the EPs and the associated Bernoulli
variance (second row) for the examples indicated in
Fig. \ref{illus_bivarDens}. The EPs increase with the correlation as
there is a strong tendency to have concurrently low temperature and
salinity. The Bernoulli variance is similarly large for high
correlations. EPs and Bernoulli variances are the same for standard
deviation $1$ or $2$, which implies that high variability in
temperature and salinity is not captured in the $p(1-p)$ expression.

\begin{table}[!h] \centering \caption{EP and Bernoulli variance for
    different correlations and variances (top rows), and expected
    Bernoulli variances for both temperature and salinity data $\by$ and 
    temperature $y_T$ (bottom rows).}
  \begin{tabular}{c|ccc|ccc}
 &\multicolumn{3}{c}{$\sigma_S=\sigma_T=1$} & \multicolumn{3}{c}{$\sigma_S=\sigma_T=2$} \\
\hline
Correlation $\gamma$ & 0.2 & 0.6 & 0.8 & 0.2 & 0.6 & 0.8 \\
\hline
$p$ & 0.28 & 0.35 & 0.40 & 0.28 & 0.35 & 0.40 \\ 
$p(1-p)$ & 0.20 & 0.23 & 0.24 & 0.20 & 0.23 & 0.24 \\ 
$E_{\by}(p(\by) (1-p(\by)))$ & 0.092 & 0.089 & 0.085 & 0.052 & 0.051 & 0.049 \\ 
$E_{y_T}(p(y_T) (1-p(y_T)))$ & 0.151 & 0.138 & 0.123 & 0.137 & 0.114 & 0.093 \\ 
\hline
\end{tabular}
\label{tab:sim_rhoab}
\end{table}

Table \ref{tab:sim_rhoab} (bottom two rows) shows results of expected
Bernoulli variance calculations. This is presented for a design
gathering both data types $(y_T,y_S)$, and for a design with
temperature measurements $y_T$ alone. With both data,
$(y_T,y_S)^t=(\xi_T,\xi_S)^t+N(0,0.5^2\bI)$, while
$y_T=\xi_T+N(0,0.5^2)$ when only temperature is measured.  For this illustration, Table \ref{tab:sim_rhoab} shows that the expected Bernoulli variance
gets lower with larger standard deviations $\sigma_T$ and $\sigma_S$
(right columns). The reduction of Bernoulli variance is largest
for the cases with high correlation $\gamma$. Albeit smaller, there is
also uncertainty reduction when only temperature is measured (bottom
row), especially when temperature and salinity are highly correlated. When correlation is low ($\gamma=0.2$), there is little
information about salinity in the temperature data, and therefore less
uncertainty reduction. In an application with fresh cold water from
river source, the temperature and salinity variables will not only be
interdependent, but will show dependence in the spatial
dimension. This in turn will impact the design criteria when we
evaluate the information measure by integrating over several locations
$\bx$.

@@

An important concept underlying this work is the potential of robotic sampling to adapt survey plans based on data and statistical inference
in accordance with the sense-plan-act loop
(Fig. \ref{fig:sense-plan-act}). We will therefore present a few adaptive strategies
for robotic sampling that build on the notion of multivariate ES. Simulations, analysis, and  results from full-scale trials are presented.

@@

An adaptive AUV survey is split into many rounds (stages). In each
round, the survey design is created using a myopic approach that considers the largest reduction in uncertainty of the ES. In our
setting, the selection is made on the graph and defined by the nearest grid
nodes in the domain $\mathcal{M}_g$.

The optimal sequential design not only considers the best current AUV
grid node, but also how the data gathered at this node could inform future sampling at successive stages. Let $d^{j,s}$ denote design
number $j$ at stage $s$ of a sequential survey. If this design is
selected, data $\by^{j,s}$ will be gathered. The optimal path
selection situation is  depicted in Figure \ref{fig:PathSelOpt},
\begin{figure}[h!]
\centering
\includegraphics[width=0.85\textwidth]{Figures/sequent_select.pdf}
\caption{Optimal sequential path selection.}\label{fig:PathSelOpt}
\end{figure}
where design choices are indicated by squares while data realizations are indicated by circles. 
%terms this optimal design is defined as
%\begin{equation}\label{opt_crit}
%    \bd^* = \mbox{argmin}_{d^{j,1}} \left\{ \int_{\by^{j,1}} \mbox{argmin}_{d^{j,2}} \left\{ \int_{\by^{j,2}} \ldots \pi(\by^{j,2}|\by^{j,1}) d\by^{j,2} \right\} \pi(\by^{j,1}) d \by^{j,1} \right\},
%\end{equation}
%where $\ldots$ represents the expected variance reduction in the ES under the continued optimal design, which will depend on the data at earlier stages. 
The mathematical expression for the optimal design then involves a
series of intermixed maximizations over designs and integrals over
data.  In practice, the optimal solution is intractable because of
the enormous growth over stages (see
e.g. \cite{sucar2015probabilistic} and \cite{powell2016perspectives}).
Instead, we outline heuristic strategies. In terms of notation, let
$\mathcal{Y}^{s-1} =\{d^{j,r}, \by^{j,r} ; r=1,\ldots,s-1 \}$ denote
the data gathered until stage $s-1$, using a selected design
$d^{j,r}$, $r=1,\ldots,s-1$.

The derived closed-form solutions for the expected IBV are still
important building blocks when constructing the sampling designs, as
they will be used to score the different adaptive designs. % Efficient
% calculation is important for using adaptive survey designs with
% robotic platforms.
